---
title: "Insights & 2027 Projections"
format:
  html:
    code-fold: true
    code-tools: true
  pdf:
    colorlinks: true
    papersize: a4
    fontsize: 11pt
    margin-left: 2cm
    margin-right: 2cm
    margin-top: 2cm
    margin-bottom: 2cm
execute:
  echo: false
  warning: false
  message: false
---

::: callout-important
## 🚀 Key Executive Takeaways

-   **Exponential Growth:** Adoption of EVs in Turkey is proceeding exponentially, not linearly. The market is accelerating faster than traditional forecasts.
-   **Investment Bottlenecks:** Based on our 2027 estimates, there will be a critical shortage of charging stations not only in urban centers but particularly in **transit-route provinces**.
-   **Actionable Prioritization:** At a benchmark of "50 EVs per station," immediate capital injection is required in the critical (red) zones mapped below to sustain Turkey's EV targets.
:::

## 1. 📊 Correlation Reality Check

Before projecting the future, we tested the current relationship between electric vehicles and charging infrastructure.

```{r setup, warning=FALSE, message=FALSE}
#| code-fold: true
#| code-summary: "Click to View Libraries & Data Prep Code"

# =============================================================================
# Turkey EV Infrastructure Analysis — 2027 Forecast & Bottleneck Mapping
# Project: Team-EV-Turkey
# =============================================================================

# ---- 1. Libraries -----------------------------------------------------------
required_pkgs <- c("tidyverse", "scales", "rnaturalearth", "rnaturalearthdata",
                   "sf", "stringi", "leaflet")
for (pkg in required_pkgs) {
  if (!require(pkg, character.only = TRUE)) {
    install.packages(pkg, repos = "[https://cloud.r-project.org](https://cloud.r-project.org)")
    library(pkg, character.only = TRUE)
  }
}

# ---- 2. Load data -----------------------------------------------------------
load("clear_data/tidy_ev_data_FINAL.RData")

# ---- 3. Helper: normalize province names for joining ------------------------
normalize_name <- function(x) {
  x |>
    stri_trans_general("Latin-ASCII") |>
    str_to_lower() |>
    str_replace_all("[^a-z]", "") |>
    str_trim()
}

# ---- 4. Build analysis-ready dataset ----------------------------------------
analysis_ready <- final_df |>
  mutate(
    ev_per_station  = ev_sayisi / istasyon_sayisi,
    ev_market_share = (ev_sayisi / otomobil_sayisi) * 100,
    date            = as.Date(paste0(str_replace(tarih, "_", "-"), "-01")),
    province_key    = normalize_name(sehir)
  ) |>
  arrange(sehir, date)

# ---- 4.5 Hotfix: source-data anomaly in November 2025 -----------------------
analysis_ready <- analysis_ready |>
  group_by(sehir) |>
  arrange(date) |>
  mutate(
    ev_sayisi = if_else(
      ev_sayisi < lag(ev_sayisi) * 0.5 & ev_sayisi < lead(ev_sayisi) * 0.5,
      round((lag(ev_sayisi) + lead(ev_sayisi)) / 2),
      ev_sayisi
    )
  ) |>
  ungroup()

# ---- 5. Correlation: EV adoption vs. station investment ---------------------
global_cor <- cor(analysis_ready$ev_sayisi,
                  analysis_ready$istasyon_sayisi,
                  use = "complete.obs")
cat("\n--- EVIDENCE ---\n")
cat("Correlation between EV count and station count:", round(global_cor, 4), "\n")
```

> **Insight:** A high correlation proves that current station investments reactively follow market demand. **Our goal is to shift this from reactive to proactive.**

## 2. 📈 Market Growth Trend (Top 5 Hubs)

When examining the time-series growth chart of the top 5 provinces, the exponential curve is undeniable.

```{r trend-plot, warning=FALSE, message=FALSE}
#| code-fold: true
#| code-summary: "Click to View Trend Analysis Code"

# ---- Latest snapshot per province -------------------------------------------
latest_snapshot <- analysis_ready |>
  group_by(sehir, province_key) |>
  filter(date == max(date)) |>
  summarise(
    current_stations = istasyon_sayisi,
    current_evs      = ev_sayisi,
    .groups = "drop"
  )

# ---- Shared theme -----------------------------------------------------------
theme_ev <- function() {
  theme_minimal(base_size = 12) +
    theme(
      plot.title       = element_text(face = "bold", size = 15),
      plot.subtitle    = element_text(color = "grey40", size = 11),
      plot.caption     = element_text(color = "grey50", size = 9),
      panel.grid.minor = element_blank(),
      legend.position  = "right"
    )
}

# ---- PLOT 1: EV growth trend in major cities --------------------------------
top_5_cities <- latest_snapshot |>
  slice_max(current_evs, n = 5) |>
  pull(sehir)

p1 <- ggplot(analysis_ready |> filter(sehir %in% top_5_cities),
             aes(x = date, y = ev_sayisi, color = sehir)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.4) +
  scale_y_continuous(labels = label_comma()) +
  scale_color_brewer(palette = "Set1", name = "Province") +
  labs(
    title    = "EV Adoption Trend in Top-5 Provinces",
    subtitle = "Monthly registered electric vehicles",
    x        = NULL,
    y        = "Registered EVs",
    caption  = "Source: EPDK"
  ) +
  theme_ev()

print(p1)
```

## 3. 🎯 2027 Projections & Investment Gaps

We forecasted the 2027 EV count using a **Log-Linear Model** and identified provincial investment gaps based on an optimal benchmark of **50 EVs per station**.

| Province Tier | Action Required by 2027 |
|:-----------------------------------|:-----------------------------------|
| 🔴 **Critical Risk** | Immediate capital injection; infrastructure is severely lagging. |
| 🟠 **High Risk** | Infrastructure expansion must double to meet forecasted demand. |
| 🟢 **Adequate** | Current growth pace is sufficient. |

```{r forecast-plot, warning=FALSE, message=FALSE}
#| code-fold: true
#| code-summary: "Click to View 2027 Forecast & Gap Analysis Code"

# ---- Forecast EVs to 2027 ---------------------------------------------------
target_date_num <- as.numeric(as.Date("2027-01-01"))

forecast_2027 <- analysis_ready |>
  group_by(sehir, province_key) |>
  group_modify(~{
    df <- .x
    if (nrow(df) < 3 || length(unique(df$ev_sayisi)) < 2) {
      return(tibble(ev_2027_linear = NA_real_,
                    ev_2027_loglin = NA_real_,
                    cagr_pct       = NA_real_))
    }
    lin <- lm(ev_sayisi ~ as.numeric(date), data = df)
    pred_lin <- as.numeric(predict(lin, newdata = data.frame(date = target_date_num)))

    if (any(df$ev_sayisi <= 0)) {
      pred_log <- NA_real_; cagr <- NA_real_
    } else {
      llm <- lm(log(ev_sayisi) ~ as.numeric(date), data = df)
      slope <- coef(llm)[2]
      pred_log <- exp(as.numeric(predict(llm, newdata = data.frame(date = target_date_num))))
      cagr <- (exp(slope * 365) - 1) * 100
    }
    tibble(ev_2027_linear = round(pmax(pred_lin, 0)),
           ev_2027_loglin = round(pmax(pred_log, 0)),
           cagr_pct       = round(cagr, 1))
  }) |>
  ungroup() |>
  mutate(ev_2027 = coalesce(ev_2027_loglin, ev_2027_linear))

# ---- Priority scoring -------------------------------------------------------
TARGET_EV_PER_STATION <- 50

priority_table <- forecast_2027 |>
  inner_join(latest_snapshot, by = c("sehir", "province_key")) |>
  mutate(
    stations_needed_2027  = ceiling(ev_2027 / TARGET_EV_PER_STATION),
    investment_gap        = pmax(stations_needed_2027 - current_stations, 0),
    bottleneck_ratio_2027 = ev_2027 / pmax(current_stations, 1),
    priority_tier = case_when(
      ev_2027 > 15000 & bottleneck_ratio_2027 > 100 ~ "Critical",
      ev_2027 > 5000  & bottleneck_ratio_2027 > 75  ~ "High",
      bottleneck_ratio_2027 > 50                    ~ "Medium",
      TRUE                                          ~ "Adequate"
    ),
    priority_tier = factor(priority_tier,
                           levels = c("Critical", "High", "Medium", "Adequate"))
  ) |>
  arrange(desc(investment_gap))

# ---- PLOT 2: Supply-vs-demand gap (dumbbell chart) --------------------------
gap_plot_data <- priority_table |>
  slice_max(investment_gap, n = 15) |>
  mutate(sehir = fct_reorder(sehir, investment_gap))

p2 <- ggplot(gap_plot_data) +
  geom_segment(aes(x = current_stations, xend = stations_needed_2027,
                   y = sehir, yend = sehir),
               color = "grey70", linewidth = 1.2) +
  geom_point(aes(x = current_stations,     y = sehir,
                 color = "Current stations"), size = 4) +
  geom_point(aes(x = stations_needed_2027, y = sehir,
                 color = "Stations needed by 2027"), size = 4) +
  scale_color_manual(
    values = c("Current stations"        = "#1f77b4",
               "Stations needed by 2027" = "#d62728"),
    name = NULL) +
  scale_x_continuous(labels = label_comma()) +
  labs(
    title    = "Charging Infrastructure Gap by 2027",
    subtitle = paste0("Top-15 provinces by investment gap  ·  benchmark = ",
                      TARGET_EV_PER_STATION, " EVs per station"),
    x        = "Number of charging stations",
    y        = NULL,
    caption  = "Forecast: log-linear model on monthly EPDK data"
  ) +
  theme_ev() +
  theme(legend.position = "top")

print(p2)
```

## 4. 🗺️ Turkey 2027: The Interactive Heatmap

This is where the investments must go. The static map shows the risk distribution, while the **Interactive Map** below calculates the exact USD required to close the gap in each province *(assuming a \$50k cost per DC fast charger)*.

::: callout-note
**Hover Instructions:** Use your mouse to explore the interactive map below to view specific 2027 EV forecasts, missing stations, and required capital per province.
:::

```{r map-plots, warning=FALSE, message=FALSE}
#| code-fold: true
#| code-summary: "Click to View Spatial & Interactive Mapping Code"

# ---- Build spatial layer ----------------------------------------------------
tr_map <- ne_states(country = "turkey", returnclass = "sf") |>
  mutate(
    province_raw = name,
    province_key = normalize_name(name)
  )

manual_fixes <- tribble(
  ~ne_key,        ~tuik_key,
  "afyon",        "afyonkarahisar",
  "kmaras",       "kahramanmaras",
  "kahramanmars", "kahramanmaras",
  "icel",         "mersin",
  "urfa",         "sanliurfa",
  "zinguldak",    "zonguldak"
)

tr_map <- tr_map |>
  left_join(manual_fixes, by = c("province_key" = "ne_key")) |>
  mutate(province_key = coalesce(tuik_key, province_key)) |>
  select(-tuik_key)

map_plot_data <- tr_map |>
  left_join(priority_table, by = "province_key")

# ---- PLOT 3: Static heat map ------------------------------------------------
map_plot_data_cap <- map_plot_data |>
  mutate(bottleneck_capped = pmin(bottleneck_ratio_2027, 200))

p3 <- ggplot(map_plot_data_cap) +
  geom_sf(aes(fill = bottleneck_capped), color = "white", linewidth = 0.15) +
  scale_fill_gradientn(
    colors   = c("#fff5eb", "#fdae61", "#d7191c"),
    values   = c(0, 0.5, 1),
    na.value = "grey85",
    name     = "EVs per station\n(2027 forecast)",
    breaks   = c(50, 100, 150, 200),
    labels   = c("50", "100", "150", "200+")
  ) +
  labs(
    title    = "Turkey 2027 Charging Infrastructure Bottleneck Map",
    subtitle = "Red provinces require urgent charging-station investment"
  ) +
  theme_void(base_size = 12) +
  theme(legend.position = "right")

print(p3)

# ---- Interactive Leaflet map ------------------------------------------------
AVG_STATION_COST_USD <- 50000

map_plot_data_interactive <- map_plot_data |>
  mutate(
    investment_cost_usd = investment_gap * AVG_STATION_COST_USD,
    cost_label = scales::dollar(investment_cost_usd, prefix = "$"),
    ev_label   = scales::comma(ev_2027, accuracy = 1)
  )

pal <- colorBin(
  palette  = "YlOrRd",
  domain   = map_plot_data_interactive$investment_gap,
  bins     = c(0, 10, 50, 100, 250, 500, Inf)
)

interactive_map <- leaflet(data = map_plot_data_interactive) |>
  addProviderTiles(providers$CartoDB.Positron) |>
  addPolygons(
    fillColor   = ~pal(investment_gap),
    weight      = 1,
    opacity     = 1,
    color       = "white",
    dashArray   = "3",
    fillOpacity = 0.8,
    highlightOptions = highlightOptions(
      weight = 3, color = "#666", dashArray = "", fillOpacity = 0.9, bringToFront = TRUE
    ),
    label = ~paste0(
      province_raw,
      " | 2027 EV Forecast: ", ev_label,
      " | Missing Stations: ", investment_gap,
      " | Required Investment: ", cost_label
    ) |> lapply(htmltools::HTML)
  ) |>
  addLegend(
    pal      = pal,
    values   = ~investment_gap,
    opacity  = 0.7,
    title    = "2027 Missing Stations<br>(Investment Opportunity)",
    position = "bottomright"
  )

print(interactive_map)
```

## References

1. **Energy Market Regulatory Authority (EPDK).** (2025). *Electric Vehicle Charging Infrastructure Report - Turkey.* Retrieved from [epdk.gov.tr](https://www.epdk.gov.tr)
2. **Turkish Statistical Institute (TUIK).** (2025). *Road Motor Vehicles Statistics, December 2024.* Retrieved from [tuik.gov.tr](https://www.tuik.gov.tr)
3. **International Energy Agency (IEA).** (2024). *Global EV Outlook 2024: Trends and Projections.*
4. **TEHAD (Turkish Electric and Hybrid Vehicles Association).** (2025). *Annual EV Market Analysis for Turkey.* Retrieved from [tehad.org](https://tehad.org)
5. **Quarto Documentation.** (2026). *Creating Interactive Documents with R.* Retrieved from [quarto.org](https://quarto.org)